Linear interaction of treatment with baseline risk was sufficient for predicting individualized benefit

### 2021-10-26 submission/manuscript.rmd
start restructuring methods - figures are not final

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in RCT simulations. **Study Design and Setting:** We simulated
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its intereraction with treatment (none, linear or quadratic). In
 each sample we predicted absolute benefit using: models with the PI and a
 constant relative treatment effect; models including an interaction of treatment
 with the PI; stratification in quarters of the PI; nonlinear transformations of
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
 strength, and the PI's discriminative ability. **Results:** Models including a
 PI by treatment interaction performed better under most simulation settings.
 Flexible models required larger sample sizes and higher AUC of the PI to
 outperform the linear interaction model. The adaptive approach tended to select
 simpler models with smaller sample sizes and/or worse prediction performance.
 **Conclusion:** Under most circumstances, a model with a linear interaction of
 the PI with treatment is the optimal risk-based approach to predicting absolute
 treatment benefit.
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[pagewise, left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

For each patient we generated 8 baseline covariates here
$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Treatment
was allocated using a 50:50 split. Outcomes for patients in the control arm were
generated from a logistic regression model including all baseline covariates. In
the base case scenario coefficient values were such, that the AUC of the
logistic regression model was $0.75$ and the event rate in the control arm was
$20\%$.

While binary outcomes in the control arm were generated from Bernoulli variables
with true probabilities $P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$, 
the outcomes in the treatment arm were based on true probabilities 
$\text{expit}(lp_1)$, with $$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0$$

The coefficients $\gamma_0,\gamma_1$ and $\gamma_2$ along with the
centering constant $c$ were set for each simulation scenario to mimic a wide
variety scenarios, ranging from true constant relative treatment effect
($\gamma_1=\gamma_2=0$) to moderate and strong linear ($\gamma_2=0$) and
quadratic deviations. We also considered scenarios with treatment-covariate
interactions.
These scenarios included 4 weak interactions
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82$), 
4 strong interactions 
($\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61$), 
and 2 weak and 2 strong interactions.

The sample size of the base case scenario was set to 4,250 ($80\%$ power for the
detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
sample sizes of 1,064 and 17,000, respectively. We also evaluated the effect of
worse or better discriminative ability for risk, adjusting the baseline
covariate coefficients, such that the AUC of the regression model in the control
arm was 0.65 and 0.85 respectively.

Combining all these settings resulted in a simulation study of 66 scenarios
(exact settings in the supplementary material).

>>>>

---
title: |
 Linear interaction of treatment with baseline risk was sufficient for predicting
 individualized benefit
abstract: |
 data using diverse assumptions for a baseline prognostic index of risk (PI) and
 the shape of its intereraction with treatment (none, linear or quadratic). In
 constant relative treatment effect; models including an interaction of treatment
 with the PI; stratification in quarters of the PI; nonlinear transformations of
 the PI (restricted cubic splines with 3, 4 and 5 knots); an adaptive approach
 using Akaike’s Information Criterion. We evaluated predictive performance using
 root mean squared error and measures of discrimination and calibration for
 benefit. Starting from a base case scenario (sample size 4,250, treatment odds
 ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment effect
 PI by treatment interaction performed better under most simulation settings.
 Flexible models required larger sample sizes and higher AUC of the PI to
 outperform the linear interaction model. The adaptive approach tended to select
 simpler models with smaller sample sizes and/or worse prediction performance.
 the PI with treatment is the optimal risk-based approach to predicting absolute
 treatment benefit.
 **Objective:** To compare different risk-based methods predicting individualized
 treatment effects in RCT simulations. **Study Design and Setting:** We simulated
 each sample we predicted absolute benefit using: models with the PI and a
 strength, and the PI's discriminative ability. **Results:** Models including a
 **Conclusion:** Under most circumstances, a model with a linear interaction of
authors:
  - name: Alexandros Rekkas
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Peter R. Rijnbeek
    department: Department of Medical Informatics
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
  - name: Ewout W. Steyerberg
    department: Department of Biomedical Data Sciences
    affiliation: Leiden University Medical Center
    location: Leiden, The Netherlands
  - name: David van Klaveren
    department: Department of Public Health
    affiliation: Erasmus Medical Center
    location: Rotterdam, The Netherlands
keywords:
  - treatment effect heterogeneity
  - absolute benefit
  - prediction models
output: 
  rticles::arxiv_article:
    keep_tex: true
  bookdown::word_document2:
    reference_docx: reference.docx
geometry: margin=1.0in
date: false
toc: false
font-size: 11pt
header-includes:
  - \renewcommand*\familydefault{\sfdefault}
  - \usepackage{setspace}
  - \usepackage{amsmath}
  - \doublespacing
  - \usepackage[pagewise, left]{lineno}
  - \usepackage{amssymb}
  - \usepackage{bm}
  - \usepackage{caption}
  - \usepackage{booktabs}
  - \date{}
  - \newcommand\given[1][]{\:#1\vert\:}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
csl: jce.csl
---

@@

For each patient we generated 8 baseline covariates here
$x_1,\dots,x_4\sim N(0, 1)$ and $x_5,\dots,x_8\sim B(1, 0.2)$. Treatment
was allocated using a 50:50 split. Outcomes for patients in the control arm were
generated from a logistic regression model including all baseline covariates. In
the base scenarios coefficient values were such, that the AUC of the
logistic regression model was $0.75$ and the event rate in the control arm was
$20\%$. Binary outcomes in the control arm were generated from Bernoulli variables
with true probabilities 
$P(y=1|X, t_x = 0) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}$.

Outcomes in the treatment arm were generated using 3 base scenarios: absent
treatment effect (OR = 1), moderate treatment effect (OR = 0.8) and high
treatment effect (OR = 0.5). We started with simulating outcomes based on true
constant relative treatment effects for the 3 base scenarios. We then simulated
linear, quadratic and non-monotonic deviations from constant treatment effects
using:
$$lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0, $$
where $lp_1$ is the true linear predictor in the treatment arm, so that
$P(y=1|X, t_x=1) = \text{expit}(lp_1)$. Finally, we simulated scenarios where a
constant absolute harm is applied across all treated patients. In this case we
have $P(y=1|X,t_x=1) = \text{expit}(lp_1) + \text{harm}$.

The sample size for the base scenarios was set to 4,250 ($80\%$ power for the
detection of a marginal OR of 0.8). We evaluated the effect of smaller or larger
sample sizes of 1,063 and 17,000, respectively. We also evaluated the effect of
worse or better discriminative ability for risk, adjusting the baseline
covariate coefficients, such that the AUC of the regression model in the control
arm was 0.65 and 0.85 respectively.

Combining all these settings resulted in a simulation study of 486 scenarios
(exact settings in the supplementary material).

### 2021-10-26 submission/manuscript.tex
start restructuring methods - figures are not final

\begin{abstract}
\textbf{Objective:} To compare different risk-based methods predicting
individualized treatment effects in RCT simulations. \textbf{Study
Design and Setting:} We simulated data using diverse assumptions for a
baseline prognostic index of risk (PI) and the shape of its
intereraction with treatment (none, linear or quadratic). In each sample
we predicted absolute benefit using: models with the PI and a constant
relative treatment effect; models including an interaction of treatment
with the PI; stratification in quarters of the PI; nonlinear
transformations of the PI (restricted cubic splines with 3, 4 and 5
knots); an adaptive approach using Akaike's Information Criterion. We
evaluated predictive performance using root mean squared error and
measures of discrimination and calibration for benefit. Starting from a
base case scenario (sample size 4,250, treatment odds ratio 0.8, AUC of
the PI 0.75), we varied the sample size, the treatment effect strength,
and the PI's discriminative ability. \textbf{Results:} Models including
a PI by treatment interaction performed better under most simulation
settings. Flexible models required larger sample sizes and higher AUC of
the PI to outperform the linear interaction model. The adaptive approach
tended to select simpler models with smaller sample sizes and/or worse
prediction performance. \textbf{Conclusion:} Under most circumstances, a
model with a linear interaction of the PI with treatment is the optimal
risk-based approach to predicting absolute treatment benefit.
\end{abstract}

@@

For each patient we generated 8 baseline covariates here
\(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
Treatment was allocated using a 50:50 split. Outcomes for patients in
the control arm were generated from a logistic regression model
including all baseline covariates. In the base case scenario coefficient
values were such, that the AUC of the logistic regression model was
\(0.75\) and the event rate in the control arm was \(20\%\).

While binary outcomes in the control arm were generated from Bernoulli
variables with true probabilities
\(P(y=1|X) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}\), the outcomes in
the treatment arm were based on true probabilities
\(\text{expit}(lp_1)\), with
\[lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0\]

The coefficients \(\gamma_0,\gamma_1\) and \(\gamma_2\) along with the
centering constant \(c\) were set for each simulation scenario to mimic
a wide variety scenarios, ranging from true constant relative treatment
effect (\(\gamma_1=\gamma_2=0\)) to moderate and strong linear
(\(\gamma_2=0\)) and quadratic deviations. We also considered scenarios
with treatment-covariate interactions. These scenarios included 4 weak
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.82\)), 4 strong
interactions (\(\text{OR}_{t_x=1} / \text{OR}_{t_x=0}=0.61\)), and 2
weak and 2 strong interactions.

The sample size of the base case scenario was set to 4,250 (\(80\%\)
power for the detection of a marginal OR of 0.8). We evaluated the
effect of smaller or larger sample sizes of 1,064 and 17,000,
respectively. We also evaluated the effect of worse or better
discriminative ability for risk, adjusting the baseline covariate
coefficients, such that the AUC of the regression model in the control
arm was 0.65 and 0.85 respectively.

Combining all these settings resulted in a simulation study of 66
scenarios (exact settings in the supplementary material).

>>>>

\begin{abstract}
data using diverse assumptions for a baseline prognostic index of risk
(PI) and the shape of its intereraction with treatment (none, linear or
quadratic). In constant relative treatment effect; models including an
interaction of treatment with the PI; stratification in quarters of the
PI; nonlinear transformations of the PI (restricted cubic splines with
3, 4 and 5 knots); an adaptive approach using Akaike's Information
Criterion. We evaluated predictive performance using root mean squared
error and measures of discrimination and calibration for benefit.
Starting from a base case scenario (sample size 4,250, treatment odds
ratio 0.8, AUC of the PI 0.75), we varied the sample size, the treatment
effect PI by treatment interaction performed better under most
simulation settings. Flexible models required larger sample sizes and
higher AUC of the PI to outperform the linear interaction model. The
adaptive approach tended to select simpler models with smaller sample
sizes and/or worse prediction performance. the PI with treatment is the
optimal risk-based approach to predicting absolute treatment benefit.
\textbf{Objective:} To compare different risk-based methods predicting
individualized treatment effects in RCT simulations. \textbf{Study
Design and Setting:} We simulated each sample we predicted absolute
benefit using: models with the PI and a strength, and the PI's
discriminative ability. \textbf{Results:} Models including a
\textbf{Conclusion:} Under most circumstances, a model with a linear
interaction of
\end{abstract}

@@

For each patient we generated 8 baseline covariates here
\(x_1,\dots,x_4\sim N(0, 1)\) and \(x_5,\dots,x_8\sim B(1, 0.2)\).
Treatment was allocated using a 50:50 split. Outcomes for patients in
the control arm were generated from a logistic regression model
including all baseline covariates. In the base scenarios coefficient
values were such, that the AUC of the logistic regression model was
\(0.75\) and the event rate in the control arm was \(20\%\). Binary
outcomes in the control arm were generated from Bernoulli variables with
true probabilities
\(P(y=1|X, t_x = 0) = \text{expit}(PI)=\frac{e^{PI}}{1+e^{PI}}\).

Outcomes in the treatment arm were generated using 3 base scenarios:
absent treatment effect (OR = 1), moderate treatment effect (OR = 0.8)
and high treatment effect (OR = 0.5). We started with simulating
outcomes based on true constant relative treatment effects for the 3
base scenarios. We then simulated linear, quadratic and non-monotonic
deviations from constant treatment effects using:
\[lp_1 = \gamma_2(PI-c)^2 + \gamma_1(PI-c) + \gamma_0, \] where \(lp_1\)
is the true linear predictor in the treatment arm, so that
\(P(y=1|X, t_x=1) = \text{expit}(lp_1)\). Finally, we simulated
scenarios where a constant absolute harm is applied across all treated
patients. In this case we have
\(P(y=1|X,t_x=1) = \text{expit}(lp_1) + \text{harm}\).

The sample size for the base scenarios was set to 4,250 (\(80\%\) power
for the detection of a marginal OR of 0.8). We evaluated the effect of
smaller or larger sample sizes of 1,063 and 17,000, respectively. We
also evaluated the effect of worse or better discriminative ability for
risk, adjusting the baseline covariate coefficients, such that the AUC
of the regression model in the control arm was 0.65 and 0.85
respectively.

Combining all these settings resulted in a simulation study of 486
scenarios (exact settings in the supplementary material).
